Practical Effects of Integrating Temperature with Strang Split Reactions

### 2021-03-08 strang/paper.tex
update refs
update

%% \author[0000-0002-5749-334X]{J.~B.~Bell}
%% \affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

%% \author[0000-0003-1791-0265]{A.~J.~Nonaka}
%% \affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

%% \author[0000-0001-8092-1974]{W.~Zhang}
%% \affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

%% \correspondingauthor{Michael Zingale}
%% \email{michael.zingale@stonybrook.edu}

@@

We are using the freely-available \castro\ simulation
code~\citep{castro,castro_joss}.  \castro\ solves the equations of
hydrodynamics using an unsplit piecewise parabolic method coupled with
gravity and reactions.  For reactive flow, either Strang splitting or
spectral deferred corrections are used to couple the hydrodynamics and
reactions~\citep{castro_sdc}.  Here we focus on the Strang splitting.

In a Strang split evolution~\cite{strang:1968}, we update the full
hydrodynamics state, $\Uc$ as:
\begin{equation}
  \Uc^{n+1} = {\bf R}_{\Delta t/2} {\bf A}_{\Delta t} {\bf R}_{\Delta t/2} \Uc^n
\end{equation}
where ${\bf R}_{\Delta t/2}$ is the reaction update through a timestep
$\Delta t/2$ and ${\bf A}_{\Delta t}$ is the advective update through
$\Delta t$.  We see with this splitting, each process operates on the
state left behind by the previous operation, and the staggering of the
physics is done to give second order accuracy in time.

@@

\item Evolve $(X_k, T)$, but ``freeze'' the specific
  heat, $c_v$, at the start of integration to avoid potentially
  expensive equation of state calls.  This was discussed in
  \cite{Bell:2004} and is the default method in \castro.

\item Evolve $(X_k, T)$, calling the equation of state each time
  we evaluate the righthand side of the system to get an updated
  $c_v$.  This is an option in \castro.

\item Evolve $(X_k, e)$, and get $T$ from $e$ using the equation of
  state.
\end{itemize}
Depending on how vigorous the burning is and how much the temperature
changes during a hydrodynamic timestep, one or more of these
approximations may be reasonable.  For explosive reactions, we expect
that evolving the full system will be required.  The goal of this
note is to try to quantify the convergence of a reacting hydrodynamics
code with these different approximations.

In \citet{castro_sdc}, we introduced a test problem where we could
measure the convergence of a reacting hydrodynamics problem via
Richardson extrapolation\MarginPar{ref?}---this was an acoustic pulse with helium
burning via $3$-$\alpha$ and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$.  Those tests focused on
SDC coupling and showed that we can get overall 4th order in space and
time convergence.  Here we run the same test with Strang coupling,
looking at the various approaches at incorporating a temperature
equation in the reactive portion of the update.

For each method, we run the \castro\ {\tt reacting\_convergence} test
problem at 5 resolutions: $64^2$, $128^2$, $256^2$, $512^2$, and
$1024^2$, with the timestep kept fixed in proportion to the grid resolution.
We then compute four errors between adjacent resolutions by
coarsening the finer resolution run, and computing the $L_1$ norm over
all zones.  Figure~\ref{thefigure} shows the norm of the error
vs.\ the coarse run resolution.  Guidelines showing ideal first and
second order convergence are shown.

\begin{figure}[t]
\centering
\plotone{test}
\caption{\label{thefigure} our results}
\end{figure}

@@

Looking at global convergence of a reacting hydrodynamics problem we
see that second order convergence is only realized when temperature is
evolved alongside the composition when using an operator-split
approach to reactions.  This is just a single, rather simple problem,
but this suggests that reactive hydrodynamics simulations should
switch to integrating temperature or another energy equation together
with reactions to yield better overall convergence and accuracy.  We
expect that for problems with vigorous burning, such as detonation,s
this will be especially important.

>>>>

%% \author[0000-0002-1530-781X]{A.\ Harpole}
%% \affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
%%              Stony Brook, NY 11794-3800}

%% \author[0000-0003-2300-5165]{D.\ E.\ Willcox}
%% \affiliation{Lawrence Berkeley National Laboratory, Berkeley, CA}

@@

We use the freely-available \castro\ simulation
code~\citep{castro,castro_joss} to solve the equations of
hydrodynamics, using an unsplit piecewise parabolic method coupled with
reactions.  \castro\ uses either Strang splitting or
spectral deferred corrections (SDC) to couple the hydrodynamics and
reactions~\citep{castro_sdc}.  Here we focus on the Strang splitting.

In a Strang split evolution~\citep{strang:1968}, we update the full
hydrodynamics state, $\Uc$, as:
\begin{equation}
  \Uc^{n+1} = {\bf R}_{\Delta t/2} {\bf A}_{\Delta t} {\bf R}_{\Delta t/2} \Uc^n
\end{equation}
where ${\bf R}_{\Delta t/2}$ is the reaction update through a timestep
$\Delta t/2$ and ${\bf A}_{\Delta t}$ is the advective update through
$\Delta t$.  We see with this splitting, each process operates on the
state left behind by the previous operation, and the staggering of the
physics is done to give second order accuracy in time.

@@

\item Evolve $(X_k, T)$.  To avoid expensive equation of state calls
  in getting the specific heat, we can optionally ``freeze'' the value
  of $c_v$ at the start of integration to avoid potentially expensive
  equation of state calls.  This was discussed in \cite{Bell:2004} and
  until recently was the default method in \castro.

\item Evolve $(X_k, e)$, and get $T$ from $e$ using the equation of
  state.
\end{itemize}
For all of these cases, density is constant during the reaction operation.
Depending on how vigorous the burning is and how much the temperature
changes during a hydrodynamic timestep, one or more of these
methods may be reasonable.  For explosive reactions, we expect
that evolving the full system will be required.  The goal of this
note is to try to quantify the convergence of a reacting hydrodynamics
code with these different approximations.

In \citet{castro_sdc}, we introduced a test problem where we could
measure the convergence of a reacting hydrodynamics problem via
Richardson extrapolation---this was an acoustic pulse with helium
burning via $3$-$\alpha$ and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$.  Initially, the domain is
pure $\isotm{He}{4}$, but both $\isotm{C}{12}$ and $\isotm{O}{16}$ are
created as time evolves.  The published tests focused on SDC coupling
and showed that we can get overall 4th order in space and time
convergence.  Here we run the same test with Strang coupling, looking
at the various approaches at incorporating a temperature / energy equation in
the reactive portion of the update.

For each method, we run the \castro\ {\tt reacting\_convergence} test
problem at 5 resolutions: $64^2$, $128^2$, $256^2$, $512^2$, and
$1024^2$, with the timestep kept fixed in proportion to the grid resolution.
We then compute four errors between adjacent resolutions by
coarsening the finer resolution run, and computing the $L_1$ norm over
all zones.

Figure~\ref{thefigure} shows the norm of the error vs.\ the coarse run
resolution.  The slope of these lines is a measure of the convergence
rate~\citep{ORVV}.  Guidelines show ideal first and second order
convergence.  We see that all methods converge at least second order
for density, but for the thermodynamic quantities, $(\rho e)$ and $T$,
the method where only $X_k$ is evolved during reactions has larger
errors and much worse convergence than the other methods.  Looking at
the trace nuclei generated in the burning, $\isotm{C}{12}$ and
$\isotm{O}{16}$, we see a large difference between the two methods
that evolve some sort of energy and the one method where only $X_k$ is
evolved---the latter converging essentially first order at high
resolution.

\begin{figure}[t]
\centering
\plotone{test}
\caption{\label{thefigure} Convergence of fluid quantities as a function of resolution
for 3 different Strang equation systems: just evolving $X_k$, evolving $(X_k, T)$, but fixing $c_v$ at the start of integration (``frozen''), and evolving $(X_k, e)$.}
\end{figure}

@@

Looking at global convergence of a reacting hydrodynamics problem we
see that second order convergence is only realized when temperature or
energy is evolved alongside the composition when using a Strang-split
approach to reactions.  This is just a single, rather simple problem,
but this suggests that reactive hydrodynamics simulations should
switch to integrating temperature or another energy equation together
with reactions to yield better overall convergence and accuracy.  This
complements the work of \citep{muller:1986} which showed that when
evolving near nuclear statistical equilibrium, that evolving entropy
with the system is needed for stability.  We expect that for problems
with vigorous burning, such as detonations, this will be especially
important.
